---
title: "mapBees_USA_20240224"
author: "Josee Rousseau"
date: "2024-02-24"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
  html_notebook:
    fig_caption: no
    toc: yes
    toc_depth: 3
    toc_float: yes
always_allow_html: yes
---


Program name:     mapBees_USA_20240224.rmd 

Program location: ./BioD_Pollination/Programs/BeeData_Conservation_paper/Species_Summaries

Program goal:     Create maps and an histogram of bee sample sizes in USA

Last modified:		February, 2024, by Josee Rousseau


```{r libraries, echo = FALSE, warning = FALSE, message = FALSE}

library(here)
library(spData)
library(maps)
library(purrr)
library(sf)
library(dplyr)
library(viridis)
library(colorBlindness)
library(mapview)
library(ggplot2)
library(scales)
library(ggpattern)
library(stringr)
library(gridExtra)

### Import filtered bee data from SCAN and GBIF
bees <- read.csv(here("Data", "bees_Chesshire2021_USA_20231011.csv")) # 1,991,840

```
&nbsp;     
 
#### Define region and hexagon grid
```{r}

### Extract USA country
region <- st_transform(us_states, crs = 4326)
# plot(region[2])

### Dissolve the polygons as one polygon
regionDiss <- region %>% 
	 st_union() %>%
	 st_sf()

### Create hexagonal grid covering USA
apothem <- 25000 # From center to middle of a side (radius but right angle with side); in meters

createHexaGrid <- function(apothem, locPolygon) {

	loc <- st_transform(locPolygon, crs = 2163) 
	loc <- st_buffer(loc, dist = 0) # Incase there are some problems with the polygons 

	### Remove borders across states or other boundaries. Important for small states, otherwise large hexagons will never be full size.
	locDiss <- loc %>% 
	  st_union() %>%
	  st_sf()
	#mapview(locDiss)

	### Create a box and associated range around the polygon
	boxDistr <- st_bbox(loc)
	xrange <- boxDistr$xmax - boxDistr$xmin
	yrange <- boxDistr$ymax - boxDistr$ymin

	### Select size for hexagon and make grid accordingly	
	grid_spacing <- apothem * 2  # size of hexagon, in units of the CRS (i.e. meters for 5514)
	numGridCells <- c(xrange/grid_spacing , yrange/grid_spacing) %>% # number of polygons necessary
  	floor() #ceiling() # rounded up to nearest integer.
	hexa <- st_make_grid(loc, square = F, n = numGridCells)
	#mapview(hexa)

	### Clip the grid to the polygon
	polyGrid <- hexa %>% 
  	st_intersection(loc) %>% 
  	st_sf() %>%
 		mutate(hID = row_number())
	#plot(polyGrid[2])
	#print(mapview(polyGrid))
	
  ### Project back WGS84 
	hexaGrid_WGS84 <- st_transform(polyGrid, crs = 4326)
	
	return(hexaGrid_WGS84)
	
}

hexaGrid <- map2(.x = apothem, .f = createHexaGrid, .y = regionDiss)
hexaGrid.sf <- hexaGrid[[1]]

```
&nbsp;     
 
#### Create a spatial object from the bee dataframe and associate with grid cells 
```{r}

### Make sure the lat / long info are kept
bees$Long <- bees$finalLongitude
bees$Lat <- bees$finalLatitude

bees.sf <- st_as_sf(x = bees,
           coords = c("Long", "Lat"),
           crs = 4326)

### Intersect the bees with the hexagon grid
bees.grid <- sf::st_join(bees.sf, hexaGrid.sf, join = st_intersects)

```

&nbsp;     
 
#### Calculate number of bees and number of species per hexagon for past 10 and all years across USA
```{r}

### Save spatial object as dataframe
bees.grid.df <- bees.grid
st_geometry(bees.grid.df) <- NULL

### Calculate the number of bees and species per hexagon grid cell
beesPerCell.10yrs <- bees.grid.df %>%
  filter(year >= 2012 & year <= 2021) %>%
  group_by(hID) %>%
  summarise(nBees = n(), nSpp = n_distinct(scientificName), .groups = "drop")
# 2966 cells

beesPerCell.TTL <- bees.grid.df %>%
  group_by(hID) %>%
  summarise(nBees = n(), nSpp = n_distinct(scientificName), .groups = "drop")
# 3,642 cells

### Merge the results with the grid to plot and fill in NAs by zero (no bees)
bees.grid.10yrs <- merge(hexaGrid.sf, beesPerCell.10yrs, by = "hID", all.x = TRUE)
bees.grid.10yrs[is.na(bees.grid.10yrs)] <- 0
bees.grid.TTL <- merge(hexaGrid.sf, beesPerCell.TTL, by = "hID", all.x = TRUE)
bees.grid.TTL[is.na(bees.grid.TTL)] <- 0

### Calculate number (percentage) of cells with specific values
pctCells.value0.10yrs <- dim(bees.grid.10yrs[which(bees.grid.10yrs$nBees == 0),])[1] / dim(bees.grid.10yrs)[1] * 100 # 23.1%
pctCells.value100.10yrs <- dim(bees.grid.10yrs[which(bees.grid.10yrs$nBees < 100),])[1] / dim(bees.grid.10yrs)[1] * 100 # 85.6%
pctCells.value0.TTL <- dim(bees.grid.TTL[which(bees.grid.TTL$nBees == 0),])[1] / dim(bees.grid.TTL)[1] * 100 # 5.6%
pctCells.value100.TTL <- dim(bees.grid.TTL[which(bees.grid.TTL$nBees < 100),])[1] / dim(bees.grid.TTL)[1] * 100 # 57.4%

```
&nbsp;     
 
#### Add protocol and effort columns representing those piece of info
```{r}
### Create a new column called "trapType" to standardize the trap information provided
bees.grid.df$trapType <- NA
pans <- c("pan", "Pan", "bowl", "Bowl", "Cup", "cup")
bees.grid.df$trapType <- ifelse(str_detect(bees.grid.df$samplingProtocol, paste(pans, collapse = "|")), "pan", bees.grid.df$trapType)
nets <- c("net", "Net", "sweep", "Sweep", "swept", "dweep")
bees.grid.df$trapType <- ifelse(str_detect(bees.grid.df$samplingProtocol, paste(nets, collapse = "|")), "net", bees.grid.df$trapType)
bees.grid.df$trapType <- ifelse(str_detect(bees.grid.df$samplingProtocol, "net") & str_detect(bees.grid.df$samplingProtocol, "trap"), "net_trap", bees.grid.df$trapType)
vanes <- c("vane", "Vane")
bees.grid.df$trapType <- ifelse(str_detect(bees.grid.df$samplingProtocol, paste(vanes, collapse = "|")), "vane", bees.grid.df$trapType)
lights <- c("light", "Light")
bees.grid.df$trapType <- ifelse(str_detect(bees.grid.df$samplingProtocol, paste(lights, collapse = "|")), "light", bees.grid.df$trapType)
pits <- c("pit", "Pit", "PIT")
bees.grid.df$trapType <- ifelse(str_detect(bees.grid.df$samplingProtocol, paste(pits, collapse = "|")), "pitfall", bees.grid.df$trapType)
malaises <- c("malaise", "Malaise", "maliase", "malise", "malasie")
bees.grid.df$trapType <- ifelse(str_detect(bees.grid.df$samplingProtocol, paste(malaises, collapse = "|")), "malaise", bees.grid.df$trapType)
funnels <- c("funnel", "Funnel")
bees.grid.df$trapType <- ifelse(str_detect(bees.grid.df$samplingProtocol, paste(funnels, collapse = "|")), "funnel", bees.grid.df$trapType)
nests <- c("Nest", "nest")
bees.grid.df$trapType <- ifelse(str_detect(bees.grid.df$samplingProtocol, paste(nests, collapse = "|")), "nest", bees.grid.df$trapType)
vaccums <- c("Vac", "VAC", "vac", "suction trap")
bees.grid.df$trapType <- ifelse(str_detect(bees.grid.df$samplingProtocol, paste(vaccums, collapse = "|")), "vaccum", bees.grid.df$trapType)
flights <- c("flight", "Flight", "flright")
bees.grid.df$trapType <- ifelse(str_detect(bees.grid.df$samplingProtocol, paste(flights, collapse = "|")), "flight", bees.grid.df$trapType)
others <- c("Canopy Trap", "Mercury Vapor Lamp", "mercury vapor lamp", "Sticky Trap", "sticky trap", "yellowsticky trap", "Hand", "hand", "pheromone trap", "Pheromone","window trap", "Berlese Extraction", "Beat Sheet", "CO2-baited ABC trap", "SLAM trap", "Upper tier box trap", "cage trap", "beating")
bees.grid.df$trapType <- ifelse(str_detect(bees.grid.df$samplingProtocol, paste(others, collapse = "|")), "otherTraps", bees.grid.df$trapType)
bees.grid.df$trapType[is.na(bees.grid.df$trapType)] <- "unknown"

### Assess number of records per trap type
bees.grid.df.trapType <- bees.grid.df %>%
  group_by(trapType) %>%
  summarise(n = n(), .groups = "drop")

### Assess the samplingProtocol that did not get associated to a trapType
# bees.grid.df.noTrap <- bees.grid.df %>%
#   filter(trapType == "unknown") %>%
#   group_by(samplingProtocol) %>%
#   summarise(n = n(), .groups = "drop")

### Have a look at sampling effort information available
# as.data.frame(table(bees.grid.df$samplingEffort)) #  

### Add what type of effort (number of traps/observers and/or duration) is associated with the record
bees.grid.df$effortType <- NA
bees.grid.df <- mutate(bees.grid.df,
       effortType = case_when(
         str_detect(samplingEffort, "bowl") ~ "NumTraps",
         str_detect(samplingEffort, "minutes") ~ "Duration",
         # str_detect(samplingEffort, "Minutes & observers") ~ "Duration_NumTraps",
         TRUE ~ effortType))

### Assess number of records per trap type
bees.grid.df.effortType <- bees.grid.df %>%
  group_by(effortType) %>%
  summarise(n = n(), .groups = "drop")

### Assign effort type "unknown" to all records without information
bees.grid.df$effortType[is.na(bees.grid.df$effortType)] <- "unknown"

```
&nbsp;     
 
#### Histogram of the number of records by year and determine categories and colors for graph
```{r}

### Histogram of the number of records per year
ggplot(subset(bees.grid.df, year >= 1992 & year <= 2021), aes(x = year)) +
  geom_histogram(position = "identity", alpha = 0.75, binwidth = 1, stat = "count", fill = "chocolate3") +
  ylab("Number of record per year") +
  theme_bw()

ggplot(subset(bees.grid.df, year >= 1992 & year <= 2021 & trapType != "unknown"), aes(x = year)) +
  geom_histogram(position = "identity", alpha = 0.75, binwidth = 1, stat = "count", fill = "chocolate3") +
  ylab("Number of record per year - with trap") +
  theme_bw()

### Categorize the records as having protocol and effort, protocol only, or no protocol or effort (no records have effort and no protocol)
bees.grid.df$quality <- ifelse(bees.grid.df$trapType != "unknown" & bees.grid.df$effortType != "unknown", "Protocol and effort",
                       ifelse(bees.grid.df$trapType != "unknown" & bees.grid.df$effortType == "unknown", "Protocol only",
                       ifelse(bees.grid.df$institutionCode == "iNaturalist", "iNaturalist", "No protocol or effort")))

bees.grid.df$quality <- factor(bees.grid.df$quality,
                               levels = c("No protocol or effort", "iNaturalist", "Protocol only", "Protocol and effort"))

### Select color palette
bar.colors2 <- c("black", "black", "grey", "red")

```

&nbsp;     
 
#### Create histogram
```{r}

grid.arrange(ggplot(subset(bees.grid.df, year >= 1922 & year <= 2021), aes(fill = quality, x = year, pattern = quality)) +
  geom_bar_pattern(position = "stack", 
                   pattern_fill = NA, #"black",
                   pattern_color = "white",
                   pattern_angle = 45,
                   pattern_density = 0.025, # Approx. fraction of area the pattern fills
                   # pattern_frequency = 5,
                   pattern_size = 0.1, # stroke linewidth
                   linewidth = 2,  # white lines between stacked bars
                   pattern_spacing = 0.01) + # Spacing between repetitions of pattern
  scale_fill_manual(values = bar.colors2) +
  scale_pattern_manual(name = ,
                       values = c("No protocol or effort" = "none", 
                                  "iNaturalist" = "stripe", 
                                  "Protocol only" = "none", 
                                  "Protocol and effort" = "none")) +
  geom_segment(x = 1997, y = 42000,
               xend = 2018, yend = 3000,
               lineend = 'square', linejoin = 'mitre',
               linewith = 1.5,
               arrow = arrow(length = unit(0.3, "cm")), # , type = "closed"
               color = "red", show.legend = FALSE) +
  scale_y_continuous(label = comma) +  # ,  label = function(l) {trans = l / 1000; paste0(trans, "K")}
  scale_x_continuous(breaks = seq(1922, 2022, 10)) +
  ylab("Number of records per year") +
  xlab("Year") +
  theme_bw() +  # theme_bw()
  theme(axis.text = element_text(size = 7),
        axis.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.position = "bottom",
        legend.key.size = unit(4,"mm"),
        legend.margin = margin(2, 2, 2, 2),
        legend.box.margin = margin(-7, -7, -7, -7),
        # legend.justification = "right",
        legend.title = element_blank(),
        panel.grid.major = element_line(colour = "grey90", linewidth = 0.2),  # 
        panel.grid.minor = element_line(colour = "grey90", linewidth = 0.2)) +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)), bottom = "Figure 1. Number of bee records per year, for 1922 to 2021 (few records exist\nprior to 1922; n = 73,695), where the black bar section represents the number of\nrecords without protocol (trap type) nor effort (duration and/or number of\ntraps/volunteers), gray representsrecords with only protocol information, and red represents complete\nrecords containing both protocol and effort information.")

### For FEE publication
ggsave("Fig1_histNumBeeRecords_20240224.tiff", path = here("Results"), units = "mm", width = 180, height = 90, dpi = 300) 

```

&nbsp;     
 
#### Some background information about iNaturalist records
```{r}

iNat <- bees.grid.df[which(bees.grid.df$institutionCode == "iNaturalist"), c("institutionCode", "year", "samplingProtocol", "samplingEffort", "trapType")] # 175,503 records

### Percentage of all records from iNaturalist 
round(175503 / 1991840 * 100, 2) # 8.81%

### Total number of bee records per year
table(bees.grid.df$year)
#  1700  1770  1800  1805  1812  1815  1819  1822  1824  1834  1848  1853  1858  1860  1862  1864  1865  1866  1867 
#    59    12     4     1     1     1     1     1     1     1     2     1     1    15     1     8    28     1     3 
#    
#  1870  1871  1872  1873  1874  1875  1876  1877  1878  1879  1880  1881  1882  1883  1884  1885  1886  1887  1888 
#    10     3    45     1     5     4     3    54    12    36    32    23    40    52    39   418   501   372   566 
#    
#  1889  1890  1891  1892  1893  1894  1895  1896  1897  1898  1899  1900  1901  1902  1903  1904  1905  1906  1907 
#   281   393   902   683  1183  1186   970   898  1996   947  1015  1070  1020  2020  1141  3003  1442  2633  1995 
#   
#  1908  1909  1910  1911  1912  1913  1914  1915  1916  1917  1918  1919  1920  1921  1922  1923  1924  1925  1926 
#  1667  2607  2040  1992  2297  4615  3884  3641  3740  5417  2949  5138  3922  2650  3516  2810  3299  4417  3874 
#  
#  1927  1928  1929  1930  1931  1932  1933  1934  1935  1936  1937  1938  1939  1940  1941  1942  1943  1944  1945 
#  4425  3294  5066  5042  4589  3681  4811  7942  6225  9454  9638  7347  6097  5944  5065  3538  1475  2578  1532 
#  
#  1946  1947  1948  1949  1950  1951  1952  1953  1954  1955  1956  1957  1958  1959  1960  1961  1962  1963  1964 
#  5086  5023  6915 14245 15680 10897 19576 12395 14334 11555 16056  9252 12965 14242 14260 16717 19860 16778 20998 
#  
#  1965  1966  1967  1968  1969  1970  1971  1972  1973  1974  1975  1976  1977  1978  1979  1980  1981  1982  1983 
# 18081 19406 19651 11143 15366 16824 14986 19512 11638 13550 11967 12574 13022 10654 10726 13431 13757 11383 14505 
# 
#  1984  1985  1986  1987  1988  1989  1990  1991  1992  1993  1994  1995  1996  1997  1998  1999  2000  2001  2002 
# 10047  8616  7419  7094  5912  7005  5406  6348  7860  8841  9672 12274  9505 16320 10525  9303 12446 16541 21433 
# 
#  2003  2004  2005  2006  2007  2008  2009  2010  2011  2012  2013  2014  2015  2016  2017  2018  2019  2020  2021 
# 24767 19923 27116 25987 28612 31455 34370 59050 71764 54008 37142 37381 64571 37042 35410 43580 35119 51628 68964 

table(iNat$year)
#  1929  1940  1943  1946  1950  1957  1959  1960  1966  1969  1972  1978  1981  1983  1984  1987  1988  1991  1996 
#     1     1     2     1     1     1     3     1     1     2     1     1     1     2     1     2     3     2     1 

#  1997  1999  2000  2002  2003  2004  2005  2006  2007  2008  2009  2010  2011  2012  2013  2014  2015  2016  2017 
#     2     1     4     3     7    16    36    26    51    41   127   205   242   503   903  1328  1881  4640  7665 

#  2018  2019  2020  2021 
# 14595 27442 51607 64150 

### For 2019, 2020, 2021, what is the percentage of the records consisting of iNat
iNat.2019.pct <- 27442 / 35119 * 100 # 78.14 %
iNat.2020.pct <- 51607 / 51628 * 100 # 99.96 %
iNat.2021.pct <- 64150 / 68964 * 100 # 93.02 %

### Total number of records from iNat for years 2019 to 2021
ttlNumRecords.2019to2021 <- 35119+51628+68964
numRecordsFromiNat.2019to2021 <- 27442+51607+64150
numRecordsFromiNat.2019to2021/ttlNumRecords.2019to2021*100 # 91.96%

### Values present in samplingProtocol field for iNaturalist records
table(iNat$samplingProtocol)
#        
# 175503   i.e., no record had samplingProtocol information, no record had NA, they were all ""

table(iNat$samplingEffort)
#       
# 62382 i.e., no record had samplingEffort information, they were either NA or ""

```
